---
title: "MZES SSDL - Shiny Apps: Development and Deployment"
author: "Konstantin Gavras"
institute: "MZES, University of Mannheim"
date: "15 October 2019"
output: 
  beamer_presentation:
    incremental: no
    fig_caption: false
header-includes:
     - \usepackage{longtable}
     - \usepackage{xcolor}
     - \newcommand{\bsmall}{\begin{small}}
     - \newcommand{\esmall}{\end{small}}
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
img1_path <- "shiny.png"
img2_path <- "shiny-inputs.png"

library(knitr) 
```

# Shiny

## Shiny from RStudio
```{r, out.width = "50%", fig.align="center"}
include_graphics(img1_path)
```

- Shiny is a package by RStudio to build interactive web pages...
    - without having any knowledge of web development (HTML/CSS/JavaScript)

- Shiny Apps interact with R 
    - Allows for calculations, display of R objects, presentation of results ... 

## Shiny App Components

1. Front end

- the web page actually shown to the user
- the HTML page written by Shiny
- includes layout, appearance, design features


- in Shiny terminology: `ui` (user interface)

2. Back end

- code running the app, including all functions, data import, etc.
- involves the logic of the app
- responsible for creating objects on the front end

- in Shiny terminology: `server`

## Setting up a Shiny App

Shiny Apps can be set up in two different ways:

1. Single file App

- `ui` and `server` are stored in one script

- used when developing very simple Shiny Apps

- name of the file has to be `app.R`!!!

2. Two file App

- `ui` and `server` are stored in separate scripts

- clear separation between front end and back end

- highly preferable when developing more advanced Shiny Apps

- names of the files have to be `ui.R` and `server.R`!!!

$\rightarrow$ We are going to develop Shiny Apps using the Two File method

# Developing Shiny Apps - Step by Step

## Let's get started!

Workshop materials: [https://github.com/KostaGav/shiny-development-deployment](https://github.com/KostaGav/shiny-development-deployment)

**Features covered in the workshop:**

Development:

1. Building a Shiny App from scratch 

2. Building the plain UI

3. Getting output objects and control widgets into the UI

4. Implementing the server logic

5. Output/Input Reaction

6. Rendering objects

7. Reactivity

Deployment:

1. Deploy your app using `shinyapps.io`

2. Deploy on your own VM using `Shiny Server`


## Building a Shiny App from scratch

```{r, eval=FALSE, echo = TRUE}
install.packages("shiny")
library(shiny)
runExample("01_hello")
#To show alternative Apps, please type runExample(NA) 
#and choose another example
```

## Building a Shiny App from scratch

- Create a new folder with two R scripts:

`ui.R`:

```{r, eval=FALSE, echo=TRUE}
library(shiny)
ui <- fluidPage()
```

`server.R`:

```{r, eval=FALSE, echo=TRUE}
server <- function(input, output){}
```

- Launch the Shiny App by pressing the 'Run App' button in the top right corner


## Building the plain UI

- When building a Shiny App, one should have, in general, in mind how the app should 'look' like

- Thus, we build the UI first

- In simple Shiny App, the whole UI fits in the `fluidpage`

    - every new object is passed comma-separated 
    - text can be passsed to the UI by entering strings 

- In order to format text, Shiny uses HTML wrappers:

    - these wrappers are functions taking one object as argument (+ further style options)
    - `h1()`: Top-level header
    - `h2()`: secondary header
    - `strong()`: make text bold
    - `em()`: make text italicized
    - `br()`: add line break
    

    
- We can add an official header using `titlePanel()`

Q: Can you see any particular differences between using `h1()` and the `titlePanel()` when using them as title?


## Building the plain UI

- Until now, we only have a plain white page

- We need a proper layout to make it appear nicer:

    - `sidebarlayout` is the simplest layout format
    - Input and control widgets on the left side, results and plots on the right hand side
    
```{r, eval = F, echo=T}
ui <- fluidPage(
  titlePanel("Title of my Shiny App"),
  sidebarLayout(
    sidebarPanel("My input goes here"),
    mainPanel("The results go here")
  )
)
```

## Adding Input and Control Widgets

- In order to interact with Shiny Apps, we need control widgets

- User can specify inputs, enter text or select specific dates to create a certain results

- all input function have two arguments: `inputId` and `label`

    - `inputId`: name Shiny uses to refer to this input, when retrieving values for the back end
    - has to be unique! (WARNING: if you provide two Ids with the same name, there won't be an error message!)
    - `label`: Text displaying the label of the control widget
    
## Adding Input and Control Widgets

```{r, out.width = "100%", fig.align="center"}
include_graphics(img2_path)
```

## Adding Input and Control Widgets

- control widgets go in the `sidebarPanel`

- always choose control widgets depending on the design of your app!

- Most common:

    - `radioButtons()`
    - `selectInput()`
    - `sliderInput()`


## Adding Input and Control Widgets - Radio Buttons

- We specify the possible values, range and appearance in the control widget 

```{r, eval = F, echo=T}
radioButtons(
  "buttons",
  "Push a button",
  choices = c("One", "Two", "Three"),
  selected = "One"
)
```


## Adding Input and Control Widgets - Select Input

- We specify the possible values, range and appearance in the control widget 

```{r, eval = F, echo=T}
selectInput(
  "selector",
  "Select a color",
  choices = c("Blue", "Green", "Yellow")
)
```


## Adding Input and Control Widgets - Slider Input

- We specify the possible values, range and appearance in the control widget 

```{r, eval = F, echo=T}
sliderInput(
  "slider",
  "Pick a range",
  min = 0,
  max = 100,
  value = c(10,30),
  pre = "€"
)
```