---
title: "MZES SSDL - Shiny Apps: Development and Deployment"
author: "Konstantin Gavras"
institute: "MZES, University of Mannheim"
date: "15 October 2019"
output: 
  beamer_presentation:
    incremental: no
    fig_caption: false
header-includes:
     - \usepackage{longtable}
     - \usepackage{xcolor}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
img1_path <- "shiny.png"

library(knitr) 
```

# Shiny

## Shiny from RStudio
```{r, out.width = "50%", fig.align="center"}
include_graphics(img1_path)
```

- Shiny is a package by RStudio to build interactive web pages...
    - without having any knowledge of web development (HTML/CSS/JavaScript)

- Shiny Apps interact with R 
    - Allows for calculations, display of R objects, presentation of results ... 

## Shiny App Components

1. Front end

- the web page actually shown to the user
- the HTML page written by Shiny
- includes layout, appearance, design features


- in Shiny terminology: `ui` (user interface)

2. Back end

- code running the app, including all functions, data import, etc.
- involves the logic of the app
- responsible for creating objects on the front end

- in Shiny terminology: `server`

## Setting up a Shiny App
