---
title: "MZES SSDL - Shiny Apps: Development and Deployment"
author: "Konstantin Gavras"
institute: "MZES, University of Mannheim"
date: "15 October 2019"
output: 
  beamer_presentation:
    incremental: no
    fig_caption: false
header-includes:
     - \usepackage{longtable}
     - \usepackage{xcolor}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
img1_path <- "shiny.png"

library(knitr) 
```

# Shiny

## Shiny from RStudio
```{r, out.width = "50%", fig.align="center"}
include_graphics(img1_path)
```

- Shiny is a package by RStudio to build interactive web pages...
    - without having any knowledge of web development (HTML/CSS/JavaScript)

- Shiny Apps interact with R 
    - Allows for calculations, display of R objects, presentation of results ... 

## Shiny App Components

1. Front end

- the web page actually shown to the user
- the HTML page written by Shiny
- includes layout, appearance, design features


- in Shiny terminology: `ui` (user interface)

2. Back end

- code running the app, including all functions, data import, etc.
- involves the logic of the app
- responsible for creating objects on the front end

- in Shiny terminology: `server`

## Setting up a Shiny App

Shiny Apps can be set up in two different ways:

1. Single file App

- `ui` and `server` are stored in one script

- used when developing very simple Shiny Apps

- name of the file has to be `app.R`!!!

2. Two file App

- `ui` and `server` are stored in separate scripts

- clear separation between front end and back end

- highly preferable when developing more advanced Shiny Apps

- names of the files have to be `ui.R` and `server.R`!!!

$\rightarrow$ We are going to develop Shiny Apps using the Two File method

# Developing Shiny Apps - Step by Step

## Let's get started!

Workshop materials: [https://github.com/KostaGav/shiny-development-deployment](https://github.com/KostaGav/shiny-development-deployment)

**Features covered in the workshop:**

Development:

1. Building a Shiny App from scratch

2. Building the plain UI

3. Getting output objects and control widgets into the UI

4. Implementing the server logic

5. Output/Input Reaction

6. Rendering objects

7. Reactivity

Deployment:

1. Deploy your app using `shinyapps.io`

2. Deploy on your own VM using `Shiny Server`


## Building a Shiny App from scratch

```{r, eval=FALSE, echo = TRUE}
install.packages("shiny")
library(shiny)
runExample("01_hello")
#To show alternative Apps, please type runExample(NA) 
#and choose another example
```

## Building a Shiny App from scratch

- Create a new folder with two R scripts:

`ui.R`:

```{r, eval=FALSE, echo=TRUE}
library(shiny)
ui <- fluidPage()
```

`server.R`:

```{r, eval=FALSE, echo=TRUE}
server <- function(input, output){}
```

- Launch the Shiny App by pressing the 'Run App' button in the top right corner


## Building the plain UI

XXX
